Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2020-06-10 Response-to-Reviewers-v2.Rmd

---
title: "Response to Reviewers"
output: pdf_document
---

We would like to thank the editor and two anonymous reviewers for their comments and suggestions to revise the paper. In this letter we respond to the feedback received, and describe the actions taken to revise the paper.

\newpage

#Reviewer 1

This article uses spatial point pattern methods to explore the distribution of violent crime as it relates to the location of three types of churches and other land use types in Recife, Brazil. The authors suggest that the intensity of violent crimes is higher near to evangelical churches than other types of churches.

>\textcolor{blue}{Thank you for reading the paper and for your thoughtful suggestions to improve it.}

The article as currently written is lacking a thorough and logical argument as to why the location of churches can be hypothesized to influence violent crime at the point-level (i.e., where distance is the measure of interest rather than location within an area).

>\textcolor{blue}{Thank you for this comment. We believe this question has two parts: can we hypothesize that the location of churches influences violent crime? And if so, does this influence happen at the point level?}

>\textcolor{blue}{We believe that the first question is critical: what are the theoretical arguments for the influence of places of worship on crime? We can sketch this argument in the following way:}

>>\textcolor{blue}{1. Informal social controls can suppress crime by strengthening the capabilities of neighbors to control inappropriate behavior.}

>>\textcolor{blue}{2. Several types of community organizations, including schools, clubs, and neighborhood associations, can strengthen the capabilities of communities to exert informal social control.}

>>\textcolor{blue}{3. Of several types of community organizations, churches are among the most overtly moral, since they often have prescriptive codes of behavior.}

>>\textcolor{blue}{4. Why would the location of places of worship matter? Groff (2015) reviews several theoretical perspectives that address the role of informal social control in micro-places: these include "eyes on the street" and human territorial functioning. By affecting informal surveillance, movement governing rules, direct intervention, and the social climate at places, places of worship are hypothesized to contribute to informal social controls.}

>\textcolor{blue}{We believe that this argument is logical and sound, and churches, in fact, have been studied in the past for their potential to supress crime (Abdullah, 2018; Davignon, 2015; Furr, 2010; Traunmuller, 2011). A more novel aspect is the study of places of worship, as distinct entities from the members of the congregations: as Warner and Konkel (2019) note, this is an aspect that has received less attention in empirical and theoretical research. }

>\textcolor{blue}{The second question is related to the above: if places of worship somehow help to strengthen social control, does this happen at the level of points or for areas? In our view, Groff (2015) already explains why these investigations should matter at the micro-level. But there is also an analytical aspect to this question: what is an appropriate unit of support for the analysis? In this sense, we would note that aggregating point data to areas inevitably involves loss of information. Sometimes this situation is unavoidable (say, statistics are recorded for areas, so locational information is lost from the origin.) But when we have access to detailed geographical information, converting to area data is an unecessary loss of information. The question that we are asking is actually quite geographical: if there is evidence of a moral community effect, what is its distance decay? This question is best answered by considering the distance from criminal events to places of worship, instead of simple propinquity of criminal events and places of worship in areas.}

>\textcolor{blue}{We have edited the paper to increase its clarity.}

The current framing of the article also does not convincingly situate 'moral communities' within a geographical framework and does not make clear the hypothesized mechanisms that link the presence/absence of a church at a specific location with criminal behavior and, more specifically, the presence/absence of churches of specific religions with violent crimes. The authors should make sure not to conflate the processes linking crime with population characteristics (percent of individuals who are religious in an area), with religion more broadly (hellfire hypothesis), and with the location of a church in geographic space. This theoretical work is necessary to justify why this study is relevant and make clear how it contributes to the existing literature.

>\textcolor{blue}{Thank you for this comment. As pointed out above, the theoretical work is already discussed by Groff, among others. Although conceptually the study draws from much earlier research (e.g., Appleyard, Jacobs, Taylor), empirical work has lagged, as noted by Warner and Konkel (2019). This is where the current paper aims to make a contribution.}

The introduction of the paper does a really good job setting out the importance of reducing violent crime for achieving policy goals, but it is not clear how this research will contribute to developing, implementing, or evaluating policy. For this journal, I think it is necessary for the research to contribute to policy; so, how can police, religious officials, and other policymakers use the results of this study?

>\textcolor{blue}{The analysis of elements that can influence the occurrence of the criminal act is able to increase the range of tools that public authorities can use to face the problem, particularly in a reality as difficult as that of Brazil. The notion of informal social control allows new strategies to combat crime to be explored. In particular, we find that Catholic churches do not "project" any more moral community than pharmacies, which means that there is probably not much that can be done in terms of leveraging them for informal social control purposes. Spiritis churches have a lower relative risk at relatively short distances of less than approximately 50 m (which would be the side of a small block). Why is this? We cannot posit a causality relationship, but could this be due to superstitious beliefs associated by laypeople to this religion? And if so, could this be leveraged for policy purposes? Evangelical churches, on the other hand, have a higher relative risk than any other establishment or place of worship at distance between 20-100 m. A relevant question is whether this is crime by Evangelicals or crime towards Evangelicals. This is not something that the analysis can throw light on, but that deserves further investigation to determine whether it can lead to actionable policy recommendations.}

I currently find the logic of the analysis and the results to be confusing and I think that adding the results of more exploratory work before showing results of very specific analyses (i.e., those with SES quintiles and different church types) would help to provide a convincing argument that the results of this research are solid. For example, the authors may first examine the intensity of crime as it relates to all churches (aggregated), then to adjust for ses with all churches, and then to adjust for land use. Then, after presenting the results of the initial analysis for the aggregated churches, the results of the specific church types can be presented.

>\textcolor{blue}{Words.}

The authors should add a review of social disorganization theory and informal social control as these concepts are mentioned a number of times but not discussed in any detail. The similarities and differences between these concepts and the geographically-situated moral community concept should also be discussed.

>\textcolor{blue}{The referee's suggestion has been incorporated into the text. Two footnotes have been included that detail theories of social disorganization and the theory of social control. The focus was to present the main characteristics of these theoretical views and to relate them to the geographical aspect. In general, both views highlight the relationship between local characteristics (neighborhoods, for example), social networks and criminal behavior.}

The authors should justify their choice of method over other approaches, such as regression models, which provide a framework for quantifying the relationship between crime and churches after accounting for SES factors, land use types, and population at risk. There are a number of regression models that allow  coefficients to vary over space, which would quantify how the impact of churches on crime changes by distance (see: generalized additive models, spatially-varying coefficients, geographically weighted regression). Without this type of analysis, or at least a more thorough discussion of why this point pattern method was chosen, I am not convinced that crime is, in fact, associated with evangelical churches at close distances and that this isn't due to other factors, like evangelical churches being located in high population density areas (and so crime is due to higher population density, not the presence of a church), for example.

>\textcolor{blue}{Thank you for this comment. We agree, there are many methods that could be applied. In our analysis we use two tools:}

>\textcolor{blue}{1. The relative distribution, which is akin to a smooth estimate of the intensity as a function of distance.}

>\textcolor{blue}{2. The relative distribution function with a baseline function, which adjusts the relative distribution using other risk factors. In our case, those other factors are summarized by the SED score. The interpretation of the relative distribution with a baseline is as follows: values of rho greater than 1 indicate that the intensity of crime is higher over what the baseline explains, and values less than 1 indicate that the intensity of crime is lower than what the baseline explains.}

>\textcolor{blue}{In the previous version of the paper, our SED score was calculated using several potential confounders of crime: median income, proportion of popultion unemployed, proportion of households with a single mother, proportion of households with a young single mother, proportion of population in poverty. In the current version of the paper, we also include population density in the calculation of the SED score (thank you for the suggestion). Accordingly, the relative distribution with a baseline function is the smooth estimate of the intensity of crime after controlling for potential confounders, including population density.}

>\textcolor{blue}{We believe that this approach is informative, and since we are interested in the distribution of crime in the proximity of places of worship, it is also appropriate. Furthermore, by considering other underlying risk factors (in the form of the SED score), we believe that we have dilligently covered most reasonable confounders. We have edited the text to explain this.}

Please define the term 'neutral morality profile' and how it relates to the spatial distribution of crime around land uses. Please also clarify what 'having a spatial distribution commensurate with places of worship' means and provide evidence to support this claim.

>\textcolor{blue}{Following the referee suggestion, a footnote was added to the data section explaining clearly the highlighted terms. The idea of building a control group is to compare similar establishments about their spatial distribution, but different in relation to their neutrality profile in relation to crime, seeking to simulate the existence of a case and a control group.}

Did this study investigate the location of crime events, offender locations, or victim locations? I assume crime events, but it would be good to clarify nonetheless.

>\textcolor{blue}{In the data section, we changed the description. In the third paragraph, we use the term crime event location, which is more precise.}

For the intensity measures, what is the unit area for the results? Would it be possible to provide an intensity unit such as per mile^2 or per km^2, for example? I think that this would improve the interpretation of the results and Figures 6 and 7 in particular.

>\textcolor{blue}{Words.}

When discussing the quantitative results throughout Sections 5.3 and 5.4, please add uncertainty intervals so that readers can interpret the degree to which two quantities are, in fact, different? For example, p.16-line 49: "reaching a peak of 2.4916 * 10e-5 (95% CI: 2.1 * 10e-5, 3.2 * 10e-5)." (ANTÔNIO)

>\textcolor{blue}{Words.}

Please provide some additional text describing the equations in Section 3.2. It is not immediately clear to me how to interpret equations and how they are reflected in the results. For example, what is B(s), how is it calculated, and what does it mean for understanding the distribution of crime in relation to churches? (ANTÔNIO)

>\textcolor{blue}{Words.}

For readers (like me) who are unfamiliar with the geography of Recife, Brazil, it would be helpful if a detailed map was included in the paper that shows where the downtown is, major roads, different neighborhoods, etc. This could be included in the background of Figure 1 or in an Appendix. It would also be helpful if the authors provided some description of the patterns observed in Figure 1 and Figure 2 identifying where violent crimes (or churches, supermarkets, etc.) were clustered and where they weren't clustered.

>\textcolor{blue}{The appendix included information that allows a more in-depth characterization of the city of Recife: population distribution, income, inequality, urban facilities (education and health) and location of residential areas. We also have included a brief explanation of the requested clustering pattern in the text.}

How is the SES data, which I assume is distributed at the area-level, translated to points so it can be used in the point pattern analysis? Please describe how this was done and the limitations of doing so, if any. Related, why did the authors choose to break up SES into quintiles rather than keep it as a continuous measure or break up into other classifications?

>\textcolor{blue}{Thank you for this comment. When using quadrats in the earlier part of the analysis, the region is partitioned using quintiles; this is a requirement, since quadrats function as "containers" and need to be partitioned in discrete segments. However, when calculating the distribution functions, the continuous value of the SED score is used. We have clarified this in the text.}

What are the counts of the different types of churches, ice cream shops, pharmacies, and supermarkets? I am wondering if the frequency of each of these 'events' would influence the chi-square statistics and associated p-values shown in Table 3. Is this possible?

>\textcolor{blue}{Thank you for this comment. The chi-square statistic is affected if there are small numbers if events in some quadrats, but that was not the case for any of our events. We have included a new table with the quadrat counts.}

\newpage

#Reviewer 2

This was an interesting paper to review, challenging even.

>\textcolor{blue}{Thank you for reading the paper and for your thoughtful suggestions to improve it.}

We’ll start with a couple of minor issues

– in the quadrat analysis are the results conditioned on the quadrat size. If I understand correctly this might not be an issue with conditional quadrat counts, through the use of a covariate.  It does remind me of Openshaw early work on optimal zoning systems.  Perhaps a simple example to guide the reader might help.

>\textcolor{blue}{Words.}

- the symbolism in Figure 1, 2, and 3 are rather large.  I’m at a loss to interpret Figure 3, because as the image is built up (I have a rather slow graphics card) I see symbols overlapping.

>\textcolor{blue}{Words.}

There’s a cultural issue which I find challenging.  I think I know what Catholic and Evangelical churches may be, but this is first encounter with Spiritism. I assume that the first two are differentiated along some transubstantiation/consubstantiation divide, with the Catholics being in the former group. As for Spiritist churches I’m at a loss.  Do these follow the teaching of Allan Kardec; I presume it’s related to the Spiritualist churches found in the Anglophone countries, which go back to Swedenborg and Mesmer, and which seem to be originally motivated by trying to make contact with the dead. If so, then it’s very much a minority sect.  Perhaps the authors might like to expand on the details further for those readers who are not familiar with religious practices, and how they are undertaken in Brazil. Would I recognise the same in Cape Town, to take a city at random. Would this be a typical Spiritist church: https://goo.gl/maps/kwWe2B7didMk8vfDA

>\textcolor{blue}{We have included in the text an analysis of the religions studied in the article.}

Some description of the geography of Recife would also help. Where are residential areas, where are industrial zones, is there a university, and so on.

>\textcolor{blue}{The appendix included information that allows a more in-depth characterization of the city of Recife: population distribution, income, inequality, urban facilities (education and health) and location of residential areas.}

What’s the reason for the choice of controls – what if we take another three, such as take-away food outlets, garden centres, and clothing/footwear shops? Would we get similar of different results? You could use simulation to build a distribution of randomly chosen groups of different land uses. I’m not proposing that this should happen, but it’s a possibility.

>\textcolor{blue}{Words.}

I am reminded of Openshaw’s early work on cluster detection – work was based around trying to link clusters of leukaemia to electricity substations, and overhead transmission lines. Alan Craft was rumoured to have asked something along the lines of “how do we know it’s substations, or fish chip shops.  What if it’s telephone booths?”  Openshaw turned his approach on its head to ask where the clusters were occurring. This lead him to a nuclear waste reprocessing plant, and an urban domestic waste incineration plant. Does the possibility of turning the question to look at this from another point of view?

>\textcolor{blue}{Words.}

I’m left wondering whether this paper tells me something about spatial relationships in Recife.  The author(s) assert that the research is exploratory.  But to what extent might they apply elsewhere? Would these relationships translate to Vienna, for example?

>\textcolor{blue}{Words.}

I think that ASAP readers might be find this paper rarified.  They certainly need help with some of the cultural issues.

>\textcolor{blue}{Words.}

### 2020-06-10 Moral-Communities-and-Crime/Moral_Communities_and_Crime.Rmd
v2 Working on revisions after Edivaldo N.

# PCA is implemented in R as `prcomp`:
setores.pca <- setores.sf %>% 
  st_drop_geometry() %>%
  dplyr::select(median_income, prop_unemployed, prop_single_mother, prop_young_single_mother, prop_poverty) %>%
  prcomp(center = TRUE, scale = TRUE)

@@

pr_var[1:5]

@@

# Once that there is principal component (call it SED for socio-economic deprivation), it can be joined to the `sf` object before it can be rasterized and converted to an `im` object:

@@

This concern is not idle: the endemic malady of violent crime represents an important factor that threatens to derail sustainable development goals. Violent crime, furthermore, tends to be more acute in those regions most in need of development. On the one hand, violent crime represents an economic and social drag [@Becker2017analise]. The impacts are multifaceted, since they affect people's well-being, either through the loss of human life, mental health, and limitations in the right to public spaces [@Doran2011putting], or through disturbances in schooling and academic achievement [@Unodc2019development]. In turn, these negative effects combine to create the unfortunate conditions that tend to breed criminal behavior, thus creating a vicious cycle of economic disadvantage and violence [@Unodc2019development]. It is not surprising, given these peculiarities, that researchers have joined calls for increased attention to the study of the patterns of violent crime in Low and Middle Income Countries [LMIC; @Murray2013crime].

Added to the scenario, formal social control institutions in countries like Brazil leave something (or much) to be desired: while in practice they should inhibit criminal behavior, they suffer from deep deficiencies that end up reducing the deterrent power of the justice system. Serious institutional problems include the inefficiency of the police, the lack of national legislation, the glacial pace of judicial processes, and the weak situation of the prison system in the country [@Menezes2013spatial]. This makes it even more urgent to understand the role of _informal social controls_, i.e., the ability of community organizations such as schools, clubs, and neighborhood associations to suppress crime by strengthening the capabilities of neighbors to control inappropriate behavior [@Groff2015informal].

@@

With the above considerations in mind, the objective of this paper is to investigate whether and how places of worship correlate spatially with criminal events, or in other words, to investigate whether their signals as moral communities have a discernible spatial dimension. The case study is the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other. Dissagregated data allows us to analyze the relevant point patterns at a very high level of resolution.

@@

# Background: Theoretical Perspectives on Religion and Crime {#background}

@@

Despite its intuitive appeal, research has not provided much direct evidence for the hypothesis of hellfire [@Hoffmann2015religion, p. 1]. In a bizarre twist, even mainstream doctrines (such as absolution) may in fact encourage beliefs that neutralize the fear of _terrenal_ punishment, and thus turn out to be criminogenic [e.g., @Topalli2013god]. More deviant cases can even recast criminal behavior as a form of spiritual insurgency [@Chesnut2017devoted]. Counterexamples like this notwithstanding, researchers remain open to the possibility that "religion may still serve as a social control mechanism by encouraging conventional beliefs, monitoring behaviors, enhancing family attachments, or providing conventional activities" [@Hoffmann2015religion, p. 1]. In this way, the hypothesis of _moral communities_ [@Stark1996religion] recognizes that social integration is essential for increasing social control, thus reducing the practice of behavior that is not in accordance with current norms. Rohrbaugh and Jessor [-@Rohrbaugh1975religiosity] also emphasize that religion acts as social control since it defines what is an appropriate attitude according to moral values, thus making moral communities a close relative of the concept of _informal social control_, defined as “the ability of social groups or institutions to make norms or rules effective” [@Reiss1951delinquency, p. 196; cited in @Groff2015informal, p. 91].

The hypothesis of moral communities has over the years been used to examine a variety of outcomes of interest. Recent examples include Stroope and Baker's [-@Stroope2018moral] exploration of religiosity and self-rated health; Davignon and Thomson [-@Davignon2015christian] with their research on institutional context and the religiosity of students; and the study of religion as a source of trust of Traunmuller [-@Traunmuller2011moral]. This is in addition, of course, to numerous studies on criminal behavior such as Eitle [-@Eitle2011religion], an author who explored the deterrence power of religion on gambling; Lee and Bartkowski's [-@Lee2004love] investigation of juvenile homicide in rural areas; and the research of Regnerus [-@Regnerus2003moral] on adolescent delinquency. The hypothesis of moral communities as a form of informal social control, on the other hand, has been less studied from a geographical perspective, and it is only recently that has attracted the attention of researchers. Groff [-@Groff2015informal], for instance, discusses informal social control as a phenomenon that can plausibly operate at different geographical scales, from the level of the home and family, through the street block and neighborhood, possibly to other scales such as the county. In this way, Nie and Yang [-@Nie2019smoking] remark on a recent paper on the lack of research conducted to study how the religious context of a geographical area (e.g., a county) may influence (youth smoking) behavior (p. 2). This point is echoed by Warner and Konkel [-@Warner2019neighborhood], who moreover recommend the use of even smaller units of analysis (e.g., Census Block Groups), in an effort to reduce aggregation bias [see @Hipp2007block], but more importantly because the processes defined by social disorganization theory are thought to occur at relatively small geographical scales.

@@

@@

The study is of the city of Recife, the capital of the state of Pernambuco in the Northeast region of Brazil. With a population of 1,550,390 million inhabitants in 2010, Recife is one of the main Brazilian metropolises, exerting a great economic influence in neighboring regions. However, the city experiences a serious problem with violent crime, and has the dubious honor of being one of the five capital cities in Brazil with the highest homicide rates in the period under study [@Menezes2013spatial].

@@

The data are organized at the individual level and it is possible to obtain information about the location, day of the week, day of the month, period of the day, as well as gender, age, and race of the victim. The database used in this study comprises the period from July 1, 2008 to June 30, 2010. Some descriptive statistics regarding this dataset are reported in Table 1, where it can be seen that overall, about 91% of the victims of LIVC in the period analyzed were men, while approximately 98% of violent crimes were homicides. In addition, most of the victims were black or brown, and the youth population between the ages of 18 and 30 is the most affected by violent crime. Lastly, it should be noted that about 88% of the criminal events under analysis involved firearms. Figure \ref{fig:plot-crime} shows the spatial distribution of the 1,657 LIVC crimes that occurred in the city of Recife between July, 2008 and June, 2010.

@@

In addition to places of worship, the National Register of Addresses for Statistical Purposes was queried to extract facilities other than places of worship. As discussed above, the idea is to identify points of reference that can be used as controls, having a neutral morality profile. For the sake of the present study, we selected pharmacies, ice cream shops, and bakeries to construct our control group. These three types of establishments comply with the criteria of being morally neutral and having a spatial distribution commensurate with places of worship.

Figure \ref{fig:plot-events} shows the spatial distribution of the places of worhsip and control establishments in the city of Recife. Note the similarity between the maps. As expected, there are differences in the number of points, but the locations of cases and controls are quite similar.

@@

kable(sed_descriptives,
      "latex",
      booktabs = TRUE,
      digits = 2,
      escape = FALSE,
      align = c("l", "c", "c", "c", "c", "c", "c", "c"),
      caption = "\\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife",
      col.names = linebreak(c("Variable", 
                              "Min", 
                              "2nd Quartile", 
                              "Median", 
                              "Mean", 
                              "3rd Quartile", 
                              "Max", 
                              "PC Factor 1\n Loadings")),
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "scaled_down")) %>%
  footnote(general = c("The variable for young single mothers is for women aged 15-25",
                       "The first principal component accounts for 53.91% of the variance"))

@@

```{r instensity-dataframe, include=FALSE}

>>>>

# PCA is implemented in R as `prcomp`:
setores.pca <- setores.sf %>% 
  st_drop_geometry() %>%
  dplyr::select(pop_density, median_income, prop_unemployed, prop_single_mother, prop_young_single_mother, prop_poverty) %>%
  prcomp(center = TRUE, scale = TRUE)

@@

pr_var[1:6]

@@

# Once that there is a principal component (call it SED for socio-economic deprivation), it can be joined to the `sf` object before it can be rasterized and converted to an `im` object:

@@

This concern is not idle: violent crime tends to be more acute in those regions most in need of development. On the one hand, violent crime represents an economic and social drag [@Becker2017analise]. The impacts are multifaceted, since they affect people's well-being, either through the loss of human life, mental health, and limitations in the right to public spaces [@Doran2011putting], or through disturbances in schooling and academic achievement [@Unodc2019development]. In turn, these negative effects combine to create the unfortunate conditions that tend to breed criminal behavior, thus creating a vicious cycle of economic disadvantage and violence [@Unodc2019development]. It is not surprising, given these peculiarities, that researchers have joined calls for increased attention to the study of the patterns of violent crime in Low and Middle Income Countries [LMIC; @Murray2013crime].

Added to the scenario, formal social control institutions in countries like Brazil often leave something (or much) to be desired: while in practice they should inhibit criminal behavior, they suffer from deep deficiencies that end up eroding the deterrent power of the justice system. Serious institutional problems include the inefficiency of the police, the lack of national legislation, the glacial pace of judicial processes, and the weak situation of the prison system in the country [@Menezes2013spatial]. This makes it even more urgent to understand the role of _informal social controls_, i.e., the ability of community organizations such as schools, clubs, and neighborhood associations to suppress crime by strengthening the capabilities of neighbors to control inappropriate behavior [@Groff2015informal].

@@

With the above considerations in mind, the objective of this paper is to investigate whether and how places of worship correlate spatially with criminal events, or in other words, to investigate whether their signals as moral communities have a discernible spatial dimension. Our argument is as follows:

1. Since places of worship are associated with normative codes of behavior, can they by their presence contribute to create moral communities? The mechanisms for this are explained by Groff [@Groff2015informal] in terms of "Eyes on the Street" [see @Jacobs1961death; @Jacobs1968community; @Appleyard1980livable] and Human Territorial Functioning [@Taylor1988human; @Taylor1998crime].

2. If so, what is the distance decay of said moral community effect? Is it uniform over a certain distance or does it vary?

The case study is the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other. Dissagregated data allows us to analyze the relevant point patterns at a very high level of resolution.

@@

# Background {#background}

## Theoretical Perspectives on Religion and Crime

@@

Despite its intuitive appeal, research has not provided much direct evidence for the hypothesis of hellfire [@Hoffmann2015religion, p. 1]. In a bizarre twist, even mainstream doctrines (such as absolution) may in fact encourage beliefs that neutralize the fear of _terrenal_ punishment, and thus turn out to be criminogenic [e.g., @Topalli2013god]. More deviant cases can even recast criminal behavior as a form of spiritual insurgency [@Chesnut2017devoted]. Counterexamples like this notwithstanding, researchers remain open to the possibility that "religion may still serve as a social control mechanism by encouraging conventional beliefs, monitoring behaviors, enhancing family attachments, or providing conventional activities" [@Hoffmann2015religion, p. 1]. In this way, the hypothesis of _moral communities_ [@Stark1996religion] recognizes that social integration is essential for increasing social control, thus reducing the practice of behavior that is not in accordance with current norms. Rohrbaugh and Jessor [-@Rohrbaugh1975religiosity] also emphasize that religion acts as social control since it defines what is an appropriate attitude according to moral values, thus making moral communities a close relative of the concept of _informal social control_, defined as “the ability of social groups or institutions to make norms or rules effective”\footnote{The discussion about disorder is directly related to the idea of social control. According to this perspective, control mechanisms to contain deviant actions are considered indispensable [@Rodrigues2012medo]. This theory assumes that the deviant act is seen as normal in social organization, and coercive forces are needed to limit such impulses. This control can be exercised through formal institutions, as well as through informal mechanisms. It would be a set of norms as well as positive and negative sanctions, which are specified throughout the socialization process, which aims to establish values and build the individual's morals, disciplining society and subjecting individuals to standards and principles predefined. As the individual finds himself more and more inserted in the social context, less will be the chances of criminal behavior. From this perspective, the existence of bonds of solidarity and trust would be relevant for the effective exercise of supervision over crime in the neighborhood. Informal control is then linked to the ability of a neighborhood to establish communication and surveillance networks, interfering in cases of criminal actions and increasing security in the region.} [@Reiss1951delinquency, p. 196; cited in @Groff2015informal, p. 91].

The hypothesis of moral communities has over the years been used to examine a variety of outcomes of interest. Recent examples include Stroope and Baker's [-@Stroope2018moral] exploration of religiosity and self-rated health; Davignon and Thomson [-@Davignon2015christian] with their research on institutional context and the religiosity of students; and the study of religion as a source of trust of Traunmuller [-@Traunmuller2011moral]. This is in addition, of course, to numerous studies on criminal behavior such as Eitle [-@Eitle2011religion], an author who explored the deterrence power of religion on gambling; Lee and Bartkowski's [-@Lee2004love] investigation of juvenile homicide in rural areas; and the research of Regnerus [-@Regnerus2003moral] on adolescent delinquency. The hypothesis of moral communities as a form of informal social control, on the other hand, has been less studied from a geographical perspective, and it is only recently that has attracted the attention of researchers. Groff [-@Groff2015informal], for instance, discusses informal social control as a phenomenon that can plausibly operate at different geographical scales, from the level of the home and family, through the street block and neighborhood, possibly to other scales such as the county. In this way, Nie and Yang [-@Nie2019smoking] remark on a recent paper on the lack of research conducted to study how the religious context of a geographical area (e.g., a county) may influence (youth smoking) behavior (p. 2). This point is echoed by Warner and Konkel [-@Warner2019neighborhood], who moreover recommend the use of even smaller units of analysis (e.g., Census Block Groups), in an effort to reduce aggregation bias [see @Hipp2007block], but more importantly because the processes defined by social disorganization theory\footnote{The social disorganization theory (Shaw e McKay, 1942)  presents the social structure as an indispensable factor for the elaboration of theoretical and methodological models related to crime in urban neighborhoods. According to Villareal e Silva [-@Villarreal2006social], “neighborhoods characterized by dense social networks will experience greater trust among residents and cooperation in the enforcement of social norms against crime. Disadvantaged neighborhoods are thought to have fewer social ties among residents and lower rates of organizational participation”. The argument is that community characteristics such as poverty, ethnic heterogeneity and residential mobility disrupt the social organization of urban neighborhoods in such a way as to reduce their capacity to exercise social control [@Rodrigues2012medo]. Thus, aspects such as residential instability or low socioeconomic status would contribute to the weakening of the community organization, increasing crime rates. This theory highlights the importance of the geography of crime and the characteristics of disorder in its analysis.} are thought to occur at relatively small geographical scales.

@@

## Religion in Brazil

Brazil is a country with a Christian predominance. In the first national census in which religion was considered, in 1872, practically the entire population (99.72%) identified themselves as Catholic. In the 1940 census, when the religion question was researched again, Catholicism was still the religion with which the vast majority of the population identified: 95.2% of Brazilians declared themselves as Catholics that year.

The dominance of Catholicism in the religious life of Brazil is unsurprising given the colonial past of the country. However, the religious makeup of the population started was not completely homogeneous. Among other offerings in the market place of spiritual ideas, Spiritism stands out. The Spiritist Doctrine was founded in nineteenth-century France by Allan Kardec as a way of understanding the world and its relations with the "beyond": a curious mix of pseudo-scientific language, philosophy, and religion [@Arribas2011doutrina]. Spiritism arrived in Brazil in the middle of the previous century and since then it slowly spread in large and medium-sized cities. Today, Spiritism is the third largest religious segment in Brazil, after Catholicism and Evangelical Christianism, and shares with these religions the Christian principle of charity through pious assistance.

Another signal event in the developmento of religion in Brazil is the arrival of Pentecostal Protestant churches, such as the Christian Congregation of Brazil in 1910 and the Assembly of God in 1911. Initially, these religious denominations did not make important inroads in Brazilian society (as evidenced by the 1940 census). However, in the context of rapid industrialization and major urbanization, they contributed to the change in the religious landscape along with other religions [@deSouza2019pluralidade]: according to the 2010 census the configuration of the religious scene in Brazil had changed considerably from the previous century: 64.6% of the population identified as Catholic and 22.2% as Evangelical Christians. The acceleration of this demographic change was directly related to the inroads made by Evangelical churches: The Assembly of God, for instance, grew considerably to total more than 12.3 million adherents. This church is represented extensively in many regions of the country, following a system of loosely allied ministries. With a charismatic profile and the lack of a centralized authority, growth of this church has occurred mainly through subdivision into numerous small and autonomous churches. These churches have in common the Prosperity Gospel and an emphasis on economic entrepreneurship  [e.g., @Mora2008marketing]. They also do not shy away from partisan political action.

In general, there have been two major changes in Brazil regarding religious issues: the slow decline of Catholicism and the rapid growth of Evangelical Christianism. These changes are evident in the daily life of the average Brazilian, either through the radio or through TV programs, which give more and more space to religious doctrines. At the same time, there has been an increase in cases of religious intolerance in Brazil, generating violence acts against temples, objects of worship, and people.

@@

The study is of the city of Recife, the capital of the state of Pernambuco in the Northeast region of Brazil. With a population of 1,550,390 million inhabitants in 2010, Recife is one of the main Brazilian metropolises, exerting a great economic influence in neighboring regions. However, the city experiences a serious problem with violent crime, and has the dubious honor of being one of the five capital cities in Brazil with the highest homicide rates in the period under study\footnote{The appendix included information that allows a more in-depth characterization of the city of Recife: population distribution, income, inequality, urban facilities (education and health) and location of residential areas.} [@Menezes2013spatial].

@@

The data are organized at the individual level and it is possible to obtain information about the crime event location, day of the week, day of the month, period of the day, as well as gender, age, and race of the victim. The database used in this study comprises the period from July 1, 2008 to June 30, 2010. Some descriptive statistics regarding this dataset are reported in Table 1, where it can be seen that overall, about 91% of the victims of LIVC in the period analyzed were men, while approximately 98% of violent crimes were homicides. In addition, most of the victims were black or brown, and the youth population between the ages of 18 and 30 is the most affected by violent crime. Lastly, it should be noted that about 88% of the criminal events under analysis involved firearms. Figure \ref{fig:plot-crime} shows the spatial distribution of the 1,657 LIVC crimes that occurred in the city of Recife between July, 2008 and June, 2010.

@@

In addition to places of worship, the National Register of Addresses for Statistical Purposes was queried to extract facilities other than places of worship. To study the spatial relationship between churches and crime, a control group was built and then a comparison of the empirical distribution was carried out. Gilberston and Hayes [-@Gilbertson2012engineering] highlight that different configurations of urban space affect how people perceive the environment around them. In this perspective, churches are continually affecting the impression that people have of a region. Thus, we assume that churches have unique characteristics that may or may not inhibit the occurrence of a criminal act: they are urban constructions that are not neutral in relation to crime. On the other hand, there are urban facilities that do not provide any specific reason to justify the fact that the crime occurs closer or further from these buildings. Thus, it can be thought that any urban construction belongs to one of the following distinct situations: neutral or non-neutral to crime. Pharmacies, ice cream parlors and bakeries will be used in the construction of the control group and will be considered neutral. The comparison between the distributions of non-neutral and crime-neutral establishments forms the empirical basis of the study. The choice of establishments that make up the control was based on two criteria: (a) being neutral to crime and (b) having a spatial distribution similar to non-neutral establishments. The strategy used is to compare similar establishments about their spatial distribution, but different in relation to their neutrality profile in relation to crime, seeking to simulate the existence of a case and a control group. The case will be composed of churches, while the control will be pharmacies, ice cream parlors and bakeries.

As discussed above, the idea is to identify points of reference that can be used as controls, having a neutral morality profile. For the sake of the present study, we selected pharmacies, ice cream shops, and bakeries to construct our control group. These three types of establishments comply with the criteria of being morally neutral and having a spatial distribution commensurate with places of worship.

Figure \ref{fig:plot-events} shows the spatial distribution of the places of worhsip and control establishments in the city of Recife. Note the similarity between the maps. As expected, there are differences in the number of points, but the locations of cases and controls are quite similar: the distribution of the churches, as well as of the commercial establishments follow the same pattern, being correlated with the spatial distribution of the population (see in the appendix). Since such urban constructions respond to society's demand, they tend to be located close to population agglomerations.

@@

kable(sed_descriptives,
      "latex",
      booktabs = TRUE,
      digits = 2,
      escape = FALSE,
      align = c("l", "c", "c", "c", "c", "c", "c", "c"),
      caption = "\\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife",
      col.names = linebreak(c("Variable", 
                              "Min", 
                              "2nd Quartile", 
                              "Median", 
                              "Mean", 
                              "3rd Quartile", 
                              "Max", 
                              "PC Factor 1\n Loadings")),
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "scaled_down")) %>%
  footnote(general = c("The variable for young single mothers is for women aged 15-25",
                       "The first principal component accounts for 44.9% of the variance"))
```

@@

```{r counts-intensity-dataframe, include=FALSE}
